---
title: "FACS FAST2 - Single Clones HEK Analysis"
author: "Angela"
format:
  html:
    toc: true
    code-fold: true
    number-sections: true
---

# Setup and Environment Configuration 

This section loads the required R packages and sets up the file paths for the FACS data analysis.

## Load Packages

We use `flowCore` and `openCyto` for fundamental FACS data handling and gating, `ggcyto` and `ggplot2` for plotting, and the `tidyverse` suite (`dplyr`, `stringr`, etc.) for data manipulation.

```{r load-packages, message=FALSE, warning=FALSE}
# --- Load required packages ---
# Note: Installation steps are commented out but kept for completeness
# if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
# BiocManager::install(c("flowCore", "ggcyto", "openCyto"), update=FALSE, ask=FALSE)

library(flowCore)
library(ggcyto)
library(openCyto)
library(ggplot2)
library(dplyr)
library(tidyverse)
library(stringr)
library(patchwork)
library(gridExtra)
library(grid)
```

## Set Directories and Load Data

```{r}
# --- Set directories ---
# set working directory
setwd("/tachyon/groups/scratch/gdiss/Angela/03_Projects/07_splitFAST/splitFAST2/02_FACS_on_single_clones_HEK/Test_reproducible_script/")
getwd()

# Define a representative base directory path. Replace with your actual path.
base_dir <- "/tachyon/groups/scratch/gdiss/Angela/03_Projects/07_splitFAST/splitFAST2/02_FACS_on_single_clones_HEK"
fcs_dir <- file.path(base_dir, "AT20251107_GFP-mScarlet-BFP")

# Define and create the 'Plots' directory if it doesn't exist
plot_output_dir <- file.path(base_dir, "Test_reproducible_script/Plots")
if (!dir.exists(plot_output_dir)) dir.create(plot_output_dir) # <-- THIS CREATES THE FOLDER

fcs_files <- list.files(fcs_dir, pattern="\\.fcs$", full.names=TRUE)

# --- Load FCS files ---
flow_data_batch <- read.flowSet(
  files=fcs_files, 
  transformation=FALSE, 
  truncate_max_range=FALSE
)

print(flow_data_batch)
```
# Data Gating and Preprocessing 
This section defines the sequential gates for quality control (QC) and applies them to the data using a GatingSet object.

## Define and Apply Gates
We define gates for Live Cells (based on FSC-A and SSC-A) and Single Cells (based on SSC-W and SSC-H).


```{r}

# --- Define gates ---
# Live cell gate: Based on forward and side scatter area
liveGate <- rectangleGate(
  filterId="living", 
  "FSC-A"=c(100000,260000), 
  "SSC-A"=c(10000,200000)
)

# Single cell gate: Based on side scatter width vs height (to exclude doublets/aggregates)
singleCellGate <- rectangleGate(
  filterId="single", 
  "SSC-W"=c(50000,100000), 
  "SSC-H"=c(0,150000)
)

# --- Apply gates using GatingSet ---
gs <- GatingSet(flow_data_batch)

# Add the 'living' gate to the root population
gs_pop_add(gs, liveGate, parent="root")
recompute(gs)

# Add the 'single' gate to the 'living' population
gs_pop_add(gs, singleCellGate, parent="living")
recompute(gs)

# --- Extract single cell data ---
# Get the flowSet object containing only the events that passed both gates
fs_single <- gs_pop_get_data(gs, y="living/single")

```


## Gate Verification Plots 
### Life cells


```{r fig.width=8, fig.height=4, fig.align='center', warning=FALSE}
# Select the fifth sample from the original, ungated flowSet
sample_to_check <- flow_data_batch[[5]]

plot_hex_live_gate <- autoplot(sample_to_check, x = "FSC-A", y = "SSC-A") +
  # Set axis limits to the instrument's full scale for standardized viewing
  ggcyto_par_set(limits = "instrument") +
  # Use hexagonal binning (geom_hex) instead of individual points to visualize high density areas
  geom_hex(bins = 500) +
  # Overlay the liveGate defined earlier
  geom_gate(liveGate) +
  labs(title = paste0("QC: Live Gate Check on Sample 5 (", identifier(sample_to_check), ")")) +
  theme_bw(base_size = 14)

# Save the plot
ggsave(file.path(base_dir, "Test_reproducible_script/Plots/QC_LiveGate_FSC_SSC.png"), plot_hex_live_gate, width=8, height=8, dpi=300)

plot_hex_live_gate

```


### Single cells


```{r fig.height=4, fig.align='center', warning=FALSE}

# Select the fifth sample from the original, ungated flowSet
sample_to_check <- flow_data_batch[[5]]

# 1. Temporarily apply the Live Gate to the single sample to isolate the population of interest
fs_gated_temp <- Subset(sample_to_check, liveGate)
# 2. Plot the resulting live population (fs_gated_temp)
plot_hex_single_gate <- autoplot(fs_gated_temp, x = "SSC-W", y = "SSC-H") +
  # Set axis limits to the instrument's full scale for standardized viewing
  ggcyto_par_set(limits = "instrument") +
  # Use hexagonal binning to visualize density
  geom_hex(bins = 500) +
  # Overlay the singleCellGate defined earlier
  geom_gate(singleCellGate) +
  labs(title = paste0("QC: Single Cell Gate Check on Live Events (Sample 5)")) +
  theme_bw(base_size = 14)

# Save the plot
ggsave(file.path(base_dir,"Test_reproducible_script/Plots/QC_Sample5_HexBin_SingleGate.png"), plot_hex_single_gate, width=8, height=8, dpi=300)

plot_hex_single_gate
```


## Prepare Metadata
Extract relevant experimental information (cell line, replicate, time point) from the filenames --> always name them same (cellline, replicate, time treatment)

```{r}


# --- Prepare metadata ---
filenames <- sampleNames(fs_single)
meta <- do.call(rbind, lapply(filenames, function(x){
  parts <- strsplit(x, "_")[[1]]
  cell_line <- parts[3]
  replicate <- gsub("[^0-9]", "", parts[4])
  time <- parts[5]
  if(!grepl("min|h", time)) time <- NA
  data.frame(filename=x, cell_line, replicate, time, stringsAsFactors = FALSE)
}))
rownames(meta) <- filenames
pData(fs_single) <- meta

head(pData(fs_single))

```


# Data Transformation and Aggregation
This section converts the gated flowSet into a tidy data frame and calculates per-sample summary statistics.

## Function: Convert flowSet to Tidy Data
A function to transform the flowSet into a single, wide format data.frame suitable for ggplot2 and dplyr.




```{r}
# --- Function: Convert flowSet to tidy dataframe ---
process_fluorescence_data_wide <- function(flow_data) {
  if(is.null(pData(flow_data))) stop("flowSet has no pData / metadata")
  
  all_data <- lapply(seq_along(flow_data), function(i) {
    f <- flow_data[[i]]
    sample_name <- sampleNames(flow_data)[i]
    
    # Select and rename channels for clarity
    df <- as.data.frame(exprs(f)) %>%
      select(`GFP-A`, `DsRed-A`, `Alexa Fluor 405-A`, `SSC-A`) %>%
      rename(GFP=`GFP-A`, mScarlet=`DsRed-A`, BFP=`Alexa Fluor 405-A`, SSC=`SSC-A`)
    
    # Get metadata for the current sample
    meta <- pData(flow_data)[sample_name, , drop=FALSE]
    
    # Combine metadata with fluorescence data
    cbind(meta[rep(1,nrow(df)),], Sample=sample_name, df)
  }) %>% bind_rows()
  
  return(all_data)
}

df_wide <- process_fluorescence_data_wide(fs_single)
head(df_wide)
```

## Compute Mean Fluorescence per Sample
Calculate the mean fluorescence intensity for the key channels (GFP and BFP) for each unique sample.

```{r}
# --- Compute mean per sample ---
df_wide <- df_wide %>%
  group_by(Sample) %>%
  mutate(
    GFP_mean = mean(GFP, na.rm=TRUE),
    BFP_mean = mean(BFP, na.rm=TRUE)
  ) %>%
  ungroup()

# Check the new columns
distinct_means <- df_wide %>% 
  select(Sample, GFP_mean, BFP_mean) %>% 
  distinct()
#only means
head(distinct_means) 
#only whole data set
head(df_wide)
```
# Density Plotting and Analysis

This section defines a reusable plotting function and sets up parameters for various experiments, followed by the plot generation.

## Function: Density Plotting

A robust function for generating density plots of fluorescence channels, with customizable log scaling, color palettes, and facetting.

```{r density-plotting-function}
# --- Function: Convert flowSet to tidy dataframe ---
plot_density <- function(df, x_var, color_var, facet_var=NULL, colors=NULL, labels=NULL, title="", log_scale=TRUE, file=NULL) {
  
  # Base font size is reduced to 12 to help fit all elements into the small display area
  p <- ggplot(df, aes_string(x=x_var, color=color_var)) +
    geom_density(alpha=0.4, linewidth=1.8) +
    # Use conditional scale_x_log10
    {if(log_scale) scale_x_log10()} +
    scale_y_continuous(expand=c(0,0)) +
    # Apply custom colors and labels if provided
    {if(!is.null(colors)) scale_color_manual(values=colors, labels=labels)} +
    # Apply facetting if provided
    {if(!is.null(facet_var)) facet_wrap(as.formula(paste("~", facet_var)))} +
    labs(
      title=title, 
      x=paste(x_var,"Intensity"), 
      y="Density", 
      color=NULL, 
      fill=NULL
    ) +
    theme_light(base_size=12) + # <-- Base font size reduction
    theme(
      plot.title = element_text(size=18, face="bold", hjust=0.5), 
      axis.title = element_text(size=14, face="bold"),          
      axis.text = element_text(size=12),                        
      strip.text = element_text(size=13, face="bold"),          
      # Legend text size increased to maintain legibility despite smaller base size
      legend.title = element_text(size=17, face="bold"),        
      legend.text = element_text(size=16),                      
      legend.position="bottom",
      legend.direction="horizontal",
      legend.key.size=unit(1.0, "cm"),                          
      panel.border = element_rect(size=1.5),
      axis.ticks = element_line(size=1.2)
    ) +
    guides(color=guide_legend(nrow=2, byrow=TRUE))
  
  # This line saves the high-resolution file (15x8 inches) to the 'Plots' directory
  if(!is.null(file)) ggsave(file.path(base_dir, file), p, width=15, height=8, dpi=300)
  return(p)
}
```

## Define Plotting Experiments
A list defining the parameters for four different plotting scenarios (MYC, FRB, BFP, rapa), including data filtering, axis variables, colors, and file names.



```{r}

# --- Define all plotting experiments ---
plot_params <- list(
  MYC = list(
    # Filter using logical vector; note the use of global 'df_wide'
    filter_expr = df_wide$cell_line %in% c("HEK52","HEK54","HEK55","HEK48"),
    x_var="GFP",
    color_var="cell_line",
    facet_var="replicate",
    colors=c("HEK54"="#440154","HEK48"="#6A5ACD","HEK55"="#31688E","HEK52"="#FDE725"),
    labels=c("HEK52"="yFAST","HEK54"="MYC/MAX","HEK55"="MYCΔbHLH/MAX","HEK48"="empty/empty"),
    title="GFP Density by Cell Line (MYC Clones)",
    file="Test_reproducible_script/Plots/density_MYC.png"
  ),
  FRB = list(
    filter_expr = df_wide$cell_line %in% c("HEK52","HEK49","HEK53","HEK48") & 
                  (df_wide$cell_line != "HEK53" | df_wide$time=="4h"),
    x_var="GFP",
    color_var="cell_line",
    facet_var="replicate",
    colors=c("HEK49"="#440154", "HEK48"="#6A5ACD", "HEK53"="#31688E", "HEK52"="#FDE725"),
    labels=c("HEK52"="yFAST","HEK53"="FRB/FkBP","HEK49"="_/FkBP","HEK48"="empty/empty"),
    title="GFP Density by Cell Line (FRB/FkBP Clones)",
    file="Test_reproducible_script/Plots/density_FRB.png"
  ),
  BFP = list(
    filter_expr = df_wide$cell_line != "HEKwt",
    x_var="BFP",
    color_var="replicate",
    facet_var="cell_line",
    colors=c("1"="#440154","2"="#6A5ACD","3"="#31688E"),
    labels=c("1"="Replicate 1", "2"="Replicate 2", "3"="Replicate 3"),
    title="BFP Density by Cell Line (BFP Channel)",
    file="Test_reproducible_script/Plots/density_BFP.png"
  ),
  rapa = list(
    filter_expr = df_wide$cell_line=="HEK53",
    x_var="GFP",
    color_var="time",
    facet_var="replicate",
    colors=c("0min"="#FDE725","15min"="#31688E","1h"="#6A5ACD","4h"="#440154"),
    labels=c("0min"="0 min", "15min"="15 min", "1h"="1 hour", "4h"="4 hours"),
    title="GFP Density by Time of Rapamycin Treatment",
    file="Test_reproducible_script/Plots/density_rapa.png"
  )
)
```


## Generate and Display Plots
Loop through the defined parameters to generate and save each plot. The last plot in the loop will be displayed in the Quarto output.


```{r generate-plots, fig.width=8, fig.height=4, fig.align='center', warning=FALSE}
# --- Generate all plots ---
# All plots are saved in this loop; only the last one is displayed.
all_plots <- list()
for(exp_name in names(plot_params)) {
  params <- plot_params[[exp_name]]
  
  # Filter the data frame
  df_filtered <- df_wide %>% filter(params$filter_expr)
  
  # Generate the plot (which internally saves the file)
  p <- plot_density(
    df_filtered,
    x_var = params$x_var,
    color_var = params$color_var,
    facet_var = params$facet_var,
    colors = params$colors,
    labels = params$labels,
    title = params$title,
    file = params$file 
  )
  
  # --- CONFIRMATION MESSAGE ---
  cat(paste0("SUCCESS: Plot '", exp_name, "' saved to: ", file.path(base_dir, params$file), "\n"))
  
  all_plots[[exp_name]] <- p
}

# Display the Rapamycin plot as an example in the Quarto document
print(all_plots$rapa)
```



```{r}
```
